---
title: "Dataplane: Parquet utilities (specs, units, scaling, and metadata)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dataplane: Parquet utilities (specs, units, scaling, and metadata)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = TRUE
)
```

## Overview

This vignette introduces Dataplane's Parquet/Dataset utilities in a gradual way:

1) **Start with the simplest useful idea**: a *spec* that documents columns and units  
2) Add **optional unit attributes** (helpful in R workflows, no file I/O required)  
3) Introduce **Parquet I/O with embedded metadata**  
4) Only then introduce **scaling (integer storage columns)** and how round-tripping works  
5) Finally, show **dataset directories + a sidecar metadata file** and how to open/query them  
6) Close with **detection** and **legacy prefixes**

All examples are intentionally small and use a temporary directory.

---

## Packages used

Dataplane is designed to work cleanly with Arrow and data.table.

```{r packages}
library(data.table)
library(arrow)
library(dplyr)
library(dataplane)
```

---

## A small example dataset

For the vignette, we'll use a simple time series with a few typical weather fields.

```{r example-data}
dt <- data.table(
  time = as.POSIXct("2026-02-11 12:00:00", tz = "UTC") + 0:4 * 3600,
  temp = c(25.12, 25.44, 26.01, 26.55, 27.02),   # air temperature
  rh   = c(55, 52, 49, 46, 45),                  # relative humidity
  wind = c(2.2, 2.6, 2.9, 3.1, 3.0),              # wind speed
  lat  = 35.246,
  lon  = -81.026,
  wbgt = c(23.10, 23.50, 24.00, 24.60, 25.00)    # WBGT
)

dt
```

---

## 1) Specs: a lightweight "data contract"

A Dataplane **spec** is a structured description of your table's columns.

At minimum, it can capture:

- `concept`: what a column represents (if recognized)
- units in different systems (`metric_units`, `imperial_units`)
- `declared_units`: an optional override (when you want to force a specific unit label)

Later (not yet), the spec can also include **scaling intent** (`writer_scale`), but for now,
ignore scaling; this section is just about documentation.

### 1.1 Create a default spec

`dp_spec_default()` inspects column names and tries to match them to a small internal
concept catalog (synonyms). If it can't confidently match a column, it leaves fields as `NA`.

```{r spec-default}
sp <- dp_spec_default(dt, declared_system = "metric", include_unmatched = TRUE)
sp
```

### 1.2 Look at a minimal view of the spec

The full spec table has several columns; early on it's easier to just look at the most
important ones.

```{r spec-fields-minimal}
sp$fields[, .(column, concept, metric_units, imperial_units, declared_units)]
```

---

## 2) Editing specs: dp_set_units()

`dp_set_units()` updates the *spec*. It does not change your data.

This is useful when:

- your column name is ambiguous and the catalog guessed wrong
- you want to override units for a column that wasn't matched
- you want to set a `declared_units` override (strongest "label")

### 2.1 Override units for specific columns

```{r set-units}
sp <- dp_set_units(
  sp,
  columns        = "wind",
  metric_units   = "m/s",
  imperial_units = "mph"
)

# Example: force a declared unit label for WBGT (optional pattern)
sp <- dp_set_units(sp, columns = "wbgt", declared_units = "degC")

sp$fields[, .(column, concept, metric_units, imperial_units, declared_units)]
```

---

## 3) Unit attributes (optional): dp_tag_units() and dp_check_units()

Dataplane can optionally attach units as an attribute on each column (default attribute name: `"units"`).
This is **R-native**, lightweight, and can help prevent unit confusion in analysis.

Important distinction:

- `dp_set_units()` edits the spec (documentation)
- `dp_tag_units()` applies the spec to a data object (adds attributes to a copy of the data)

### 3.1 Tag missing unit attributes

```{r tag-units}
tag <- dp_tag_units(dt, sp, only_missing = TRUE, system = "declared", unit_attr = "units")

dt_u <- tag$dt

# Keep the audit short early on:
tag$did_tag[did_tag == TRUE, .(column, tag_units)]
```

Check a couple attributes:

```{r show-units-attr}
attr(dt_u$temp, "units")
attr(dt_u$wind, "units")
```

### 3.2 Validate unit consistency

`dp_check_units()` compares current unit attributes to what the spec expects.
It returns a report and warns by default (does not stop unless you ask it to).

```{r check-units}
audit_units <- dp_check_units(
  dt_u, sp,
  system      = "declared",
  unit_attr   = "units",
  warn_max    = 10,
  stop_on_any = FALSE
)

# Early in the vignette, just show non-OK rows:
audit_units[status != "ok"]
```

---

## 4) Parquet I/O with embedded metadata: dp_write() and dp_read()

Now that there is a spec (and optionally unit attributes), the next step is file I/O.

Core behavior:

- `dp_write()` writes a **single Parquet file** and embeds metadata into Parquet schema key/value pairs.
- `dp_read()` reads the file back and returns both:
  - the data
  - parsed metadata (`spec_dt`, `audit_dt`, plus raw key/values)

### 4.1 Write to a temporary file

```{r paths}
out_dir <- file.path(tempdir(), "dataplane_vignette")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
stopifnot(file.access(out_dir, 2) == 0)

pq_path <- file.path(out_dir, "example.parquet")
pq_path
```

Write a Parquet file with metadata.

Notes:
- `mode="validate"` runs `dp_check_units()` before writing (warning by default).
- `tag_missing_units=TRUE` is convenient: it tags unit attributes before writing.
- **No scaling yet** in this first I/O example (`scale = FALSE`).

```{r write-parquet-basic}
res_w <- dp_write(
  x = dt,
  path = pq_path,
  spec = sp,
  mode = "validate",
  scale = FALSE,
  tag_missing_units = TRUE,
  compression = "zstd"
)

# dp_write() returns invisibly; in a vignette it can help to show a small confirmation:
res_w$path
```

### 4.2 Read back (and re-attach units)

```{r read-parquet-basic}
res_r <- dp_read(
  path = pq_path,
  decode_scaled = TRUE,   # harmless here; nothing scaled yet
  keep_storage  = TRUE,
  attach_units  = "declared"
)

dt2   <- res_r$dt
meta2 <- res_r$meta

dt2
```

Confirm unit attributes were attached:

```{r read-parquet-basic-units}
attr(dt2$temp, "units")
attr(dt2$wbgt, "units")
```

---

## 5) Metadata-only reads: dp_read_meta()

Sometimes you want to inspect metadata without loading the full data.

```{r read-meta-only}
m <- dp_read_meta(pq_path, parse = TRUE)

# Show a small, human-friendly subset:
list(
  prefix = m$prefix,
  declared_system = m$kv[[m$keys$declared_system]],
  spec_version    = m$kv[[m$keys$spec_version]]
)
```

Show a compact view of the stored spec and audit tables:

```{r read-meta-only-compact}
m$spec_dt[, c('column', 'concept', 'declared_units')][]
m$audit_dt[, c('column', 'storage_column', 'writer_scale', 'did_scale', 'did_tag')][]
```

---

## 6) Scaling (integer storage columns): introduce gently

Scaling is opt-in. It is useful when you want consistent decimal precision and sometimes smaller storage.

Key idea:

- If `writer_scale = 100`, a value like `25.12` can be stored as integer `2512` in a companion column like `temp_i`.
- Dataplane can automatically create those storage columns on write (`dp_write(..., scale=TRUE)`).
- Dataplane can also decode them on read (`dp_read(..., decode_scaled=TRUE)`).

### 6.1 Record scaling intent in the spec: dp_set_scale()

This still does **not** change your data; it only changes what the spec says should happen *if scaling is enabled*.

```{r set-scale}
sp_sc <- sp
sp_sc <- dp_set_scale(sp_sc, columns = c("temp", "wbgt"), writer_scale = 100)
sp_sc <- dp_set_scale(sp_sc, columns = c("lat", "lon"),  writer_scale = 10000)

# Do not print the whole spec table; show columns with scaling intent:
sp_sc$fields[is.finite(writer_scale) & writer_scale != 0, .(column, writer_scale)]
```

### 6.2 Write with scaling enabled (creates storage columns)

```{r write-parquet-scaled}
pq_path_sc <- file.path(out_dir, "example_scaled.parquet")

res_w_sc <- dp_write(
  x = dt,
  path = pq_path_sc,
  spec = sp_sc,
  mode = "validate",
  scale = TRUE,
  suffix = "_i",
  drop_original = FALSE,
  tag_missing_units = TRUE,
  compression = "zstd"
)

res_w_sc$path
```

Read back and decode:

```{r read-parquet-scaled}
res_r_sc <- dp_read(
  path = pq_path_sc,
  decode_scaled = TRUE,
  keep_storage  = TRUE,
  attach_units  = "declared"
)

dt_sc <- res_r_sc$dt
dt_sc
```

Show just the storage columns (if present):

```{r read-parquet-scaled-storage-cols}
storage_cols <- grep("_i$", names(dt_sc), value = TRUE)
storage_cols
dt_sc[, ..storage_cols]
```

---

## 7) Dataset directories (Arrow Datasets) + sidecar metadata

A dataset is a directory of Parquet files intended for scalable, lazy queries.

Dataplane supports dataset metadata via:

- `dp_write_dataset()` to write the dataset **and** a sidecar metadata Parquet file (e.g., `dp_meta.parquet`)
- `dp_open()` to open the dataset and read sidecar metadata

### 7.1 Write a dataset + sidecar metadata

```{r dataset-write}
ds_dir <- file.path(out_dir, "example_dataset")
dir.create(ds_dir, recursive = TRUE, showWarnings = FALSE)

# Small demo dataset: Arrow Table from dt
tbl <- arrow::Table$create(dt)

# NOTE:
# - dp_write_dataset() writes the dataset itself using arrow::write_dataset()
# - scale_document=TRUE only documents intended storage columns; it does NOT transform the dataset data
res_ds <- dp_write_dataset(
  x = tbl,
  path = ds_dir,
  spec = sp_sc,
  scale_document = TRUE,
  prefix = "dp",
  sidecar_overwrite = TRUE,
  sidecar_compression = "zstd",
  write_format = "parquet"
)

res_ds$sidecar_path
list.files(ds_dir, recursive = TRUE)
```

### 7.2 Open the dataset (ignore sidecar as data)

The sidecar is a Parquet file; Arrow will treat it like a data file unless told to ignore it.
Dataplane's recommended pattern is to ignore prefixes like `dp_` so `dp_meta.parquet` is excluded.

```{r open-dataset}
opened <- dp_open(
  ds_dir,
  metadata = "sidecar",
  ignore_prefixes = c("dp_", "_dp_", "klimo_", "_klimo_")
)

opened$meta$sidecar_path
opened$meta$spec_dt[, c('column', 'declared_units')][]
```

### 7.3 Lazy query example

```{r dataset-query}
ds <- opened$ds

# Do not print everything; confirm that a few columns exist
head(names(ds), 10)

res <- ds %>%
  filter(wbgt > 24) %>%
  summarize(
    n = n(),
    wbgt_mean = mean(wbgt, na.rm = TRUE)
  ) %>%
  collect()

res
```

---

## 8) Detection: dp_detect() and dp_print_detect()

Use detection when you want a quick signal for whether a file/dataset likely contains Dataplane metadata.

```{r dp-detect-file}
d_file <- dp_detect(pq_path, mode = "detect")
dp_print_detect(d_file)
```

```{r dp-detect-dataset}
d_dir <- dp_detect(ds_dir, mode = "detect")
dp_print_detect(d_dir)
```

---

## 9) Prefixes and legacy metadata

Dataplane stores schema metadata keys under a prefix (default `dp:`).
If you have older files with a different prefix (e.g., `klimo:`), provide legacy prefixes:

```{r legacy-prefix}
res_any <- dp_read(
  path = pq_path,
  preferred_prefix = "dp",
  legacy_prefixes = c("klimo")
)

res_any$meta$prefix
```

---

## Practical guidance and patterns

### Recommended workflow (typical)

1. Build a spec: `dp_spec_default(dt, "metric")`
2. Edit units: `dp_set_units()`
3. Optionally tag/check units: `dp_tag_units()` / `dp_check_units()`
4. Write: `dp_write(..., spec = sp, mode = "validate")`
5. Read: `dp_read(..., attach_units = "declared")`

### When to use scaling

Scaling is useful when:

- you want stable decimal precision in storage
- you want consistent round-tripping across tools/languages
- you expect many millions of rows and care about storage types

It is optional; start without it, then add it when it clearly helps.

### When to use datasets + sidecar metadata

Use dataset directories when:

- you partition by time/geography/organization
- you want lazy queries (Arrow) without pulling everything into RAM
- the dataset will grow over time

The sidecar metadata gives you a stable place to read spec/audit information.

---

## Session information

```{r session-info}
sessionInfo()
```
