---
title: "Dataplane: Parquet utilities (specs, units, scaling, and metadata)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dataplane: Parquet utilities (specs, units, scaling, and metadata)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

~~~{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = TRUE
)
~~~

## Overview

This vignette shows how to use Dataplane’s Parquet utilities to:

- create and edit a **spec** that documents column meaning, units, and storage intent
- optionally attach **unit attributes** to data columns (lightweight, R-native)
- optionally apply **integer storage encoding** (e.g., store `25.12` as `2512`)
- write and read **Parquet files** with embedded metadata
- write and read **Arrow Datasets** (directories of Parquet files) with a sidecar metadata file
- quickly inspect (“detect”) whether a file/dataset likely contains Dataplane metadata

All examples are intentionally small and use a temporary directory.

---

## Packages used

Dataplane is designed to work cleanly with Arrow and data.table.

```{r packages}
library(dataplane)
library(data.table)
library(arrow)
library(dplyr)
```

> **Development note (not needed for published vignettes):**  
> If you are testing this vignette before installing the package, you can source your local R files instead of `library(dataplane)`.

---

## A small example dataset

For the vignette, we’ll use a simple time series with a few typical weather-ish fields.
Nothing here is required—Dataplane works with any tabular data.

```{r example-data}
dt <- data.table(
  time = as.POSIXct("2026-02-11 12:00:00", tz = "UTC") + 0:4 * 3600,
  temp = c(25.12, 25.44, 26.01, 26.55, 27.02),   # air temperature
  rh   = c(55, 52, 49, 46, 45),                  # relative humidity
  wind = c(2.2, 2.6, 2.9, 3.1, 3.0),              # wind speed
  lat  = 35.246,
  lon  = -81.026,
  wbgt = c(23.10, 23.50, 24.00, 24.60, 25.00)    # WBGT
)

dt
```

---

## 1) Specs: what they are and why they exist

A Dataplane spec is a structured description of your table’s columns.

At minimum, a spec can capture:

- **concept** (what a column represents)
- **metric/imperial/declared units** (documentation)
- **writer_scale** (how you *intend* to store values if using integer scaling)

The main entry point is `spec_default()`.

### 1.1 Create a default spec

`spec_default()` uses a small internal concept catalog (synonyms) to pre-fill what it can.
Anything it cannot confidently match is left blank (NA), by design.

```{r spec-default}
sp <- spec_default(dt, declared_system = "metric", include_unmatched = TRUE)
sp
```

### 1.2 Inspect the spec table

The spec object is an S3 object with a `fields` table inside.

```{r spec-fields}
sp$fields[]
```

Key columns you’ll see:

- `column`: the original column name
- `concept`: matched concept label (or NA)
- `metric_units`, `imperial_units`: documentation targets
- `declared_units`: optional override (see next section)
- `writer_scale`: optional numeric scale factor for integer storage

---

## 2) Editing specs: set_units() and set_scale()

A helpful mental model:

- `set_units()` and `set_scale()` **edit the spec** (documentation + intent).
- They do **not** change your data.

### 2.1 Override units for specific columns (set_units)

You can update metric/imperial units and also optionally force a declared unit.

- If `declared_units` is set, it takes precedence for “declared” operations.
- If `declared_units` is missing, Dataplane uses the spec’s `declared_system` to infer
  declared units from metric/imperial.

```{r set-units}
sp <- set_units(
  sp,
  columns        = c("wind"),
  metric_units   = "m/s",
  imperial_units = "mph"
)

# Force declared units for a single column (optional pattern)
sp <- set_units(sp, columns = "wbgt", declared_units = "degC")

sp$fields[]
```

### 2.2 Set scaling intent (set_scale)

Integer scaling is optional. When you set `writer_scale`, you’re stating:

> “If scaling is enabled at write time, multiply by this factor and store as integer.”

Examples:

- `writer_scale = 100` stores two decimals (e.g., `25.12` -> `2512`)
- `writer_scale = 10000` for lat/lon preserves 4 decimals as integers

```{r set-scale}
sp <- set_scale(sp, columns = c("temp", "wbgt"), writer_scale = 100)
sp <- set_scale(sp, columns = c("lat", "lon"),  writer_scale = 10000)

sp$fields[]
```

---

## 3) Unit attributes on columns: tag_units() and check_units()

Another mental model:

- `tag_units()` **applies** your spec to a data object by attaching unit attributes
  (e.g., `attr(x$temp, "units") <- "degC"`).
- `set_units()` edits the spec; `tag_units()` edits (a copy of) the data.

### 3.1 Tag missing unit attributes (tag_units)

`tag_units()` returns a **copy** of your data plus an audit table of what it did.

```{r tag-units}
tag <- tag_units(dt, sp, only_missing = TRUE, system = "declared", unit_attr = "units")

dt_u <- tag$dt
tag$did_tag[]
```

Check the attribute on a column:

```{r show-units-attr}
attr(dt_u$temp, "units")
attr(dt_u$wind, "units")
```

### 3.2 Validate unit consistency (check_units)

`check_units()` returns an audit table and warns if it sees issues.

This is intentionally forgiving by default:

- it warns (does not stop) unless you set `stop_on_any = TRUE`
- it reports missing columns, missing units, mismatches, and unexpected units

```{r check-units}
audit_units <- check_units(
  dt_u, sp,
  system      = "declared",
  unit_attr   = "units",
  warn_max    = 20,
  stop_on_any = FALSE
)

audit_units[]
```

---

## 4) Scaling: encode/decode in-memory

Scaling is opt-in. There are two separate concerns:

- **in-memory encoding/decoding**: useful for experimentation
- **on-write scaling**: handled automatically by `write_parquet(..., scale = TRUE)`

### 4.1 Encode: create integer storage columns (scale_encode)

This creates new columns (default suffix `_i`) containing the integer-scaled values.

```{r scale-encode}
enc <- scale_encode(dt, sp, suffix = "_i", drop_original = FALSE, round_fn = "round")

dt_enc    <- enc$dt
scale_map <- enc$scale_map

dt_enc
scale_map[]
```

### 4.2 Decode: reconstruct original values from storage columns (scale_decode)

`scale_decode()` needs the audit/plan table (the “map”) that identifies:

- base column name
- storage column name
- writer_scale used

```{r scale-decode}
dt_dec <- scale_decode(dt_enc, audit_dt = scale_map, keep_storage = TRUE, overwrite = FALSE)
dt_dec
```

---

## 5) Parquet I/O: write_parquet() and read_parquet()

Core workflow:

- `write_parquet()` writes data to a Parquet file and embeds metadata in schema key/value pairs
- `read_parquet()` reads data back and optionally:
  - decodes scaled storage columns
  - reattaches declared unit attributes

### 5.1 Write to a temporary file

```{r paths}
out_dir <- file.path(tempdir(), "dataplane_vignette")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

pq_path <- file.path(out_dir, "example.parquet")
pq_path
```

Now write.

Important arguments:

- `spec`: either `"metric"/"imperial"` or a spec object (preferred once you edit it)
- `mode`:
  - `"document"`: minimal checks, just write metadata
  - `"validate"`: run `check_units()` before writing (warns by default)
- `scale`: if TRUE, create integer storage columns at write time
- `tag_missing_units`: if TRUE, attach unit attributes before writing

```{r write-parquet}
res_w <- write_parquet(
  x = dt,
  path = pq_path,
  spec = sp,
  mode = "validate",
  scale = TRUE,
  suffix = "_i",
  drop_original = FALSE,
  tag_missing_units = TRUE,
  compression = "zstd"
)

# write_parquet() returns a list invisibly; show a few useful pieces:
names(res_w)
```

### 5.2 Read back (decode + attach units)

`read_parquet()` returns a list with:

- `$dt`: the data
- `$meta`: parsed metadata (including the spec table and audit table)

```{r read-parquet}
res_r <- read_parquet(
  path = pq_path,
  decode_scaled = TRUE,
  keep_storage  = TRUE,
  attach_units  = "declared"
)

dt2   <- res_r$dt
meta2 <- res_r$meta

dt2
```

Confirm units were attached:

```{r read-parquet-units}
attr(dt2$temp, "units")
attr(dt2$wbgt, "units")
```

---

## 6) Metadata-only reads: read_parquet_meta()

Sometimes you want to inspect metadata without loading data.

This is common when:

- you’re auditing a directory of Parquet files
- you want to confirm declared system / version / keys
- you want to see what scaling was applied

```{r read-meta-only}
m <- read_parquet_meta(pq_path, parse = TRUE)

m$prefix
names(m$kv)

# Spec and audit tables (if present)
m$spec_dt[]
m$audit_dt[]
```

---

## 7) Dataset directories (Arrow Datasets): write_dataset() and open_dataset()

Arrow Datasets are directory-based collections of Parquet files.

**Important practical note:**  
Dataplane stores dataset metadata in a sidecar Parquet file in the dataset directory
(e.g., `dp_meta.parquet`). By default, Arrow may try to treat that sidecar file as part
of the dataset unless you tell Arrow to ignore files with a prefix like `dp_`.

This vignette demonstrates the recommended call pattern: pass `ignore_prefixes = "dp_"`.

### 7.1 Write a dataset + sidecar metadata

```{r dataset-write}
ds_dir <- file.path(out_dir, "example_dataset")
dir.create(ds_dir, recursive = TRUE, showWarnings = FALSE)

# Create an Arrow Table and write as a dataset
tbl <- arrow::Table$create(dt)

res_ds <- write_dataset(
  x = tbl,
  path = ds_dir,
  spec = sp,
  scale_document = TRUE,   # documents intended storage columns; does not change data
  prefix = "dp",
  write_format = "parquet"
)

res_ds$sidecar_path

# Show what's on disk (useful for debugging / transparency)
list.files(ds_dir, recursive = TRUE)
```

### 7.2 Open the dataset and read metadata (recommended)

```{r open-dataset}
opened <- open_dataset(
  ds_dir,
  metadata = "sidecar",
  # IMPORTANT: ignore "dp_" so the sidecar dp_meta.parquet is NOT treated as a data file
  ignore_prefixes = c("dp_", "_dp_", "klimo_", "_klimo_")
)

opened$ds
opened$meta$sidecar_path
opened$meta$spec_dt[]
```

### 7.3 Query without collecting (Arrow + dplyr)

This section shows the common “filter + summarize” pattern without pulling full data into memory.

```{r dataset-query}
ds <- opened$ds

# ---- 1) Inspect available columns in the dataset schema ----
ds_cols <- names(ds)
cat("Dataset columns:\n")
print(ds_cols)

# Guardrail: if the dataset accidentally contains only the sidecar column, re-open with ignore_prefixes.
# (This should not trigger with the recommended open_dataset() call above, but it prevents confusing failures.)
if (length(ds_cols) == 1 && identical(ds_cols, "dp_meta")) {
  message(
    "Dataset schema contains only 'dp_meta'. This usually means the sidecar file ",
    "was included as a data file. Re-opening the dataset while ignoring 'dp_' files."
  )

  opened <- open_dataset(
    ds_dir,
    metadata = "sidecar",
    ignore_prefixes = c("dp_", "_dp_", "klimo_", "_klimo_")
  )
  ds <- opened$ds
  ds_cols <- names(ds)
  cat("Dataset columns after re-open:\n")
  print(ds_cols)
}

# ---- 2) Find WBGT robustly ----
has_wbgt   <- "wbgt" %in% ds_cols
has_wbgt_i <- "wbgt_i" %in% ds_cols

audit <- opened$meta$audit_dt
wbgt_scale   <- NA_real_
wbgt_storage <- NA_character_

if (!is.null(audit) && nrow(audit)) {
  wbgt_row <- audit[tolower(column) == "wbgt"][1]
  if (nrow(wbgt_row)) {
    wbgt_storage <- wbgt_row$storage_column
    wbgt_scale   <- suppressWarnings(as.numeric(wbgt_row$writer_scale))
  }
}

# ---- 3) Build a queryable wbgt expression ----
if (has_wbgt) {
  ds_q <- ds
  wbgt_expr <- quote(wbgt)

} else {
  storage_col <- if (!is.na(wbgt_storage) && nzchar(wbgt_storage) && wbgt_storage %in% ds_cols) {
    wbgt_storage
  } else if (has_wbgt_i) {
    "wbgt_i"
  } else {
    NA_character_
  }

  if (is.na(storage_col)) {
    stop(
      "No usable WBGT field found in the dataset.\n",
      "Expected either 'wbgt' or a storage column like 'wbgt_i'.\n",
      "Available columns were:\n",
      paste(ds_cols, collapse = ", ")
    )
  }

  scale_factor <- if (is.finite(wbgt_scale) && wbgt_scale != 0) wbgt_scale else 100
  cat(sprintf("Decoding WBGT from storage column '%s' using scale factor %g\n", storage_col, scale_factor))

  ds_q <- ds %>%
    mutate(wbgt = .data[[storage_col]] / scale_factor)

  wbgt_expr <- quote(wbgt)
}

# ---- 4) Run a lazy query, then collect the small result ----
res <- ds_q %>%
  filter(!!wbgt_expr > 24) %>%
  summarize(
    n = n(),
    wbgt_mean = mean(!!wbgt_expr, na.rm = TRUE)
  ) %>%
  collect()

res
```

---

## 8) Detection: detect() and print_detect()

```{r detect-file}
d_file <- detect(pq_path, mode = "detect")
print_detect(d_file)
```

```{r detect-dataset}
d_dir <- detect(ds_dir, mode = "detect")
print_detect(d_dir)
```

---

## 9) Prefixes and legacy metadata

Dataplane uses a prefix for schema keys (default `dp:`). If you have older files with
a different prefix (e.g., `klimo:`), you can specify:

- `preferred_prefix = "dp"`
- `legacy_prefixes = c("klimo")`

```{r legacy-prefix}
# Example call pattern (no actual legacy file needed for the vignette)
res_any <- read_parquet(
  path = pq_path,
  preferred_prefix = "dp",
  legacy_prefixes = c("klimo")
)

res_any$meta$prefix
```

---

## Practical guidance and patterns

### Recommended workflow (typical)

1. Start with `spec_default(dt, "metric")`
2. Edit: `set_units()` / `set_scale()`
3. Write: `write_parquet(..., spec = sp, mode = "validate", scale = TRUE)`
4. Read: `read_parquet(..., decode_scaled = TRUE, attach_units = "declared")`

### When to use scaling

Scaling is useful when:

- you want smaller files or more stable storage types
- you want consistent decimal precision across languages/tools
- you expect to store many millions of rows

Scaling is not necessary for every use case. It is intentionally opt-in.

### When to use datasets + sidecar metadata

Use dataset directories when:

- you partition by time, geography, or organization
- you want lazy queries via Arrow
- you expect the dataset to grow

The sidecar metadata is recommended because it provides a stable “one place to read” for spec/audit info.

---

## Session information

```{r session-info}
sessionInfo()
```
